---
title: "Risky Behaviors"
author: by Andrea Chamorro, Mariam Khan, Janani Shankaran; Georgetown University's
  McCourt School of Public Policy
date: "May 8, 2017"
output:
  html_document:
    theme: journal
    toc: yes
  pdf_document:
    toc: yes
subtitle: How do risky behaviors predict student academic achievement?
---

# Introduction

For several decades, the academic performance of students has been a major concern. Many studies have discovered that academic success has been strongly linked with health-related factors. According to the Center for Disease Control and the 2009 National Youth Risk Behavior Survey (YRBS), there is a negative association between health-risk behaviors and academic achievement among high school students. In other words, students with higher grades are less likely to engage in health-risk behaviors than students with lower grades. Similarly, students who do not engage in health risk behaviors are more likely to receive higher grades than students who engage in health-risk behaviors. It should be noted these associations do not prove causation. 

The objective of this study is to build upon the CDC research, in order to better understand how certain behaviors can impact the grades of students. These results can encourage schools to promote health and safety among students, which would help them to establish lifelong healthy behaviors. 

# Data

The City of Somerville's Youth Risk Behavior Survey is an annual student survey conducted at Somerville High School. Students at Somerville High School were surveyed every two years, from 2002 to 2014. The dataset includes a total of 8,003 student survey responses. The dataset can be accessed here: http://bit.ly/2nRvJYa. 

# Descriptive statistics

In terms of demographic characteristics, roughly 40.1% of students are white, 15.0% are black, 24.4% are Hispanic, 8.7% are Asian, and 11.8% identify as another race. In addition, 219 observations exhibit missingness in race. Approximately 52.4% of the sample is female, while the average age of students is 16.25. 

Among the risky behaviors, students are most likely to have engaged in sexual activity or consumed alcohol. Some of the variables exhibit considerable missingness, including variables related to hurting oneself, gang affiliation, altercations, and drug use.

```{r}

```

# Methodology 

## Decision tree

Unfortunately, the decision tree yielded a mean-F1 score of 1, which is extremely unlikely. It is suspected there may have been a problem with the data or under the hood. In general, the decision tree has been known to overfit predictive models. 

```{r}

```

## Random Forest

Two random forest models were analyzed in this study.  One of the model utilized complete observations and the other model imputed missing values using KNN through the VIM library. When using only complete observations, dropped to approximately 3,000 observations. This yielded a relatively high OOB error of 56.25% In contrast, when the model imputed missing values, there were roughly 7,000 observations. It should be noted that the missing values of  the dependent variable or the demographic factors were not imputed. Similarly, the OOB error was still high, at 56.7% While both models had low overall predictability power, it was discovered that age has the greatest importance in both models. Furthermore, gender and marijuana were relatively important in both models. 

```{r}

```

## Ordered Logistic Regression

Since the dependent variable had ranked categorical responses, an ordered logistic regression was conducted.  When analyzing the different models, the mean-F1 score and the pseudo R-squared values were taken into consideration. After analyzing several models, it was determine that the following approximate model, as indicated below, was optimal. Unfortunately, the model had a low mean-f1 score of 0.742487. However, the McFadden's pseudo R-squared value was 0.3381522. A value between 0.2 and 0.4 indicates that the model is a good fit.  It should be noted that pregnancy, sexual activity, in-school altercation, cigarette use, and marijuana had a negative, statistically significant association in many of the models. 

```{r}

```

# Conclusion

Overall, while risky behaviors are associated with lower grades, they are not strong predictors of grades. This may be attributed to the limitations of the data. It is more than likely that there were other factors which are stronger predictors of grades, but were not captured in the model, such as household type, family stability, and the income of parents. Furthermore, the self-reported nature of the data may have decreased the strength of predictability. While limitations do exist, more research should be conducted in this area, which can better inform schools and policymakers. 

# Application in the Real World 

The next step would be to ideally create the basis of a scoring engine. This engine could take into account of other academic, behavioral, and environmental factors which were not described in this study. Such an engine could help to support the mitigation of risky behaviors. 
