---
title: "Risky Behaviors"
author: by Andrea Chamorro, Mariam Khan, Janani Shankaran; Georgetown University's
  McCourt School of Public Policy
date: "May 8, 2017"
output:
  html_document:
    theme: journal
    toc: yes
  pdf_document:
    toc: yes
subtitle: How do risky behaviors predict student academic achievement?
---

# Introduction

For several decades, the academic performance of students has been a major concern. Many studies have discovered that academic success has been strongly linked with health-related factors. According to the Center for Disease Control and the 2009 National Youth Risk Behavior Survey (YRBS), there is a negative association between health-risk behaviors and academic achievement among high school students. In other words, students with higher grades are less likely to engage in health-risk behaviors than students with lower grades. Similarly, students who do not engage in health risk behaviors are more likely to receive higher grades than students who engage in health-risk behaviors. It should be noted these associations do not prove causation. 

The objective of this study is to build upon the CDC research, in order to better understand how certain behaviors can impact the grades of students. These results can encourage schools to promote health and safety among students, which would help them to establish lifelong healthy behaviors. 

# Data/Methodology

The City of Somerville's Youth Risk Behavior Survey is an annual student survey conducted at Somerville High School. Students at Somerville High School were surveyed every two years, from 2002 to 2014. The dataset includes a total of 8,003 student survey responses. The dataset can be accessed here: http://bit.ly/2nRvJYa. 

# Descriptive statistics

In terms of demographic characteristics, roughly 40.1% of students are white, 15.0% are black, 24.4% are Hispanic, 8.7% are Asian, and 11.8% identify as another race. In addition, 219 observations exhibit missingness in race. Furthermore, in terms of gender and age, approximately, 52.4% of the sample is female and the average age of students is 16.25.

```{r}

```

# Model Techniques

## Decision tree

```{r}

```

## Random Forrest

```{r}

```

## Ordered Logistic Regression

```{r}

```


# Analysis 

# Policy Implications 

# Conclusion

Overall, while risky behaviors are associated with lower grades, they are not strong predictors of grades. This may be attributed to the limitations of the data. It is more than likely that there were other factors which are stronger predictors of grades, but were not captured in the model, such as household type, family stability, and the income of parents. Furthermore, the self-reported nature of the data may have decreased the strength of predictability. While limitations do exist, more research should be conducted in this area, which can better inform schools and policymakers. 
